Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-10-22 paper/main.tex
restructure part according to christians notes

\paragraph{(C) Cellular}
In bacteria the MreB protein forms an actin-like cytoskeleton~\cite{Erickson2001}.
It can polymerize and form filaments which behave similar to actin microfilaments~\cite{Dersch2020}.
The effect of MreB on the shape of cells during growth was shown in more detail
by~\cite{Ursell2014}.
They found that MreB localizes specifically at regions of negative curvature and thus introduces
heterogeneous cell growth which results in a bending force during the growth of cells.
It also means that the cells are able to sense the curvature of the cell wall through cytoskeleteal
MreB filaments during growth.
\cite{Bratton2018} showed that the trans-membrane protein RodZ can modulate MreB density and
facilitates local MreB assembly.
This feedback mechanism will result in straight rod-shaped bacteria if not interfered from the
external environment.
These results were also confirmed by the studies of~\cite{Wang2010}.
In their experiments, a single bacterium was fixed at the bottom pole while an optically trapped
bead was placed on the upper pole.
They could now exert a force on the free part and thus study the response of the cell under
different conditions.
When treating cells with A22, an antibiotic which disassembles
MreB~\cite{IWAI2002,Gitai2005,Karczmarek2007,Bean2009}, a higher displacement with identical force
could be measured, thus indicating that MreB increases mechanical rigidity of the cell.

While some rod-shaped bacteria grow by inserting new material at the ends, \textit{E.Coli} and
\textit{B.Subtilis} utilize a method where new material is inserted along the cylindrical part of
the rod. The details of the division event (i.e. Z-Ring formation) are very intricate and are heavily
studied as a subject on their own~\cite{Harry2001}.
On a higher level view, the rod is split apart at a point close to the center of the cell, thus
creating two new daughter cells.
We know that the best predictor for when the division event occurs, is the length of the rod as
opposed to the age of bacterium~\cite{Robert2014}.
Furthermore, not only the growth of the whole colony but also the growth of individual bacteria is
exponential~\cite{Amir2014,Takeuchi2005}.
\textbf{
    Bitte etwas kürzen, da du hier keinen Review schreibst und dich deswegen um Prägnanz bemühen
    musst.
}

\paragraph{(CC) Cell-Cell Interactions \& (DC) Domain-Cell Interactions}
A colony of bacteria is known to form biofilms (Dunne 2002). This strategy is is beneficial for the
survival of the collective as it allows it to harness growth factors and migrate when necessary and
also plays an important role in bacterial infections in the biomedical sector~\cite{Ong1999}.
The formation of biofilms is only possible due to the attractive nature of bacterial interactions
towards each other and surfaces (Berne et al. 2018). To analyze bacterial adhesion to surfaces,
\cite{vanLoosdrecht1989} investigated the interactions of various bacteria with negatively charged
polystyrene.
They found that the interaction is characterized by a low Gibbs energy of $(2 - 3)kT$ per cell
which can be described by a secondary minimum of the DLVO-theory~\cite{Verwey1947,Derjaguin1993}
which is relevant for distances $\geq\SI{1}{\nano\meter}$.
At these distances, the DLVO potential acts as a generic adhesive potential.
Particles which stay in the secondary minimum are held together but even weak interactions are
enough to redisperse them which means that this state is reversible.
When approaching closer distances, more intricate mechanisms come into play in which the
heterogeneity of the materials in question needs to be considered.
\cite{Hori2010} described bacterial adhesion as a “two-phase process including an initial,
instantaneous and reversible physical phase (phase one) and a time-dependent and irreversible
molecular and cellular phase (phase two)”.

All these considerations become even more intricate when interactions between bacteria are considered.
Gram-negative bacteria (eg. E.Coli) express outer membrane proteins (adhesins) which allow them to
attach to various hosts (Beachey 1981, Vaca et al. 2019).
In addition, the flagella can play an important role in the direct adhesion between
cells~\cite{Haiko2013}.
Thus it is not correct to blindly apply the results for adhesion to surfaces to the interactions
between bacteria.

\textbf{[Man versteht nicht so recht, warum du das alles hier erzählst und was daraus folgt. Die Verbindung zu deinem Modell fehlt.]}

\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material either along the circular part of the
            rod or at the tip~\cite{Robert2014,Takeuchi2005}.\\
        (3) & Division &
            Rod-Shaped bacteria divide in the middle of the rod into two new agents.\\
            % Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        (4) & Process Parameters &
            We take a potential heterogeneity of the bacteria into account by \textbf{[wir sollten über diesen Punkt nochmals reden. 
            Das ist hier leicht missverständlich. Parameter sind auch kein intrazellularer Aspekt.]} \cite{Koutsoumanis2013}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (5) & Adhesion &
            Bacteria attract at longer distances and adhere to each other at shorter
            distances~\cite{Verwey1947,Trejo2013}.\\
        \bottomrule
    \end{tabularx}
    \label{table:simulation-aspects}
    \caption{
        This table lists aspects of cellular behaviour which we consider in our computational model.
        It is split into cellular aspects (C) which concern only an individual bacterium and
        interactions between cells (CC).
        These aspects are generalized in order to fit to as many types of rod-shaped bacteria as
        possible.
        These are the assumptions for the basic model that we are presenting here.
        In order for our methods to work, we are really only required to satisfy the assumption of
        our spatial representation (see section~\ref{subsec:spatial-representation}) and thus, these
        assumptions can also be loosened if required.
    }
\end{table}

To model the spatial mechanics of elongated bacteria \cite{Billaudeau2017}, we represent them as a
collection of auxiliary vertices $\{\vec{x}_i\}$ which are connected by springs in ascending order.
Furthermore, we assume that the cells are flexible described by their rigidity. A force $\vec{F}$
interacting between cellular agents determines the radius (thickness) of the rods and an attractive
component can model adhesion between cells.

>>>>

While some rod-shaped bacteria grow by inserting new material at the ends, \textit{E.Coli} and
\textit{B.Subtilis} utilize a method where new material is inserted along the cylindrical part of
the rod. The details of the division event (i.e. Z-Ring formation) are very intricate and are heavily
studied as a subject on their own~\cite{Harry2001}.
On a higher level view, the rod is split apart at a point close to the center of the cell, thus
creating two new daughter cells.
It was shown that the best predictor for when the division event occurs, is the length of the rod as
opposed to the age of bacterium~\cite{Robert2014}.
Furthermore, not only the growth of the whole colony but also the growth of individual bacteria is
exponential~\cite{Amir2014,Takeuchi2005}.
The growth rates of individual bacteria and other process parameters can be heterogeneous which
should be accounted for as well \cite{Koutsoumanis2013}.

Many bacteria organize into biofilms~\cite{Dunne2002,Ong1999} which play a key role in the survival
of the collective.
Their formation is only possible due to the attractive nature of bacterial interactions between
each other~\cite{Berne2018}.
This interaction is characterized by a long-range attraction and short-ranged
adherance~\cite{Verwey1947,Trejo2013}.
We therefore need account for this phenomenon as well.

\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material either along the circular part of the
            rod or at the tip~\cite{Robert2014,Takeuchi2005}.\\
        (3) & Division &
            Rod-Shaped bacteria divide in the middle of the rod into two new agents.\\
            % Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (4) & Adhesion &
            Bacteria attract at longer distances and adhere to each other at shorter
            distances~\cite{Verwey1947,Trejo2013}.\\
        \bottomrule
    \end{tabularx}
    \caption{
        This table lists aspects of cellular behaviour which we consider in our computational model.
        It is split into cellular aspects (C) which concern only an individual bacterium and
        interactions between cells (CC).
        These aspects are generalized in order to fit to as many types of rod-shaped bacteria as
        possible.
        These are the assumptions for the basic model that we are presenting here.
        In order for our methods to work, we are really only required to satisfy the assumption of
        our spatial representation (see section~\ref{subsec:spatial-representation}) and thus, these
        assumptions can also be loosened if required.
    }
    \label{table:simulation-aspects}
\end{table}
